---
title: "Practical D - Data generation"
author: "Gerko Vink"
date: "Statistical Programming with R"
output: html_document
---


---

#### Exercises

---

It is always wise to fix the random seed. I often use a seed value of `123`:
```{r}
set.seed(123)
```
at the top of my document. This ensures that all calculations in my documents are exactly reproducible. This is because the random number generator in `R` will take its origin from the specified seed value. Every subsequent random process advance the random generation by one step. When we specify a seed value we can replicate the exact chain of random processes. This is an extremely useful tool.

When an `R` instance is started, there is initially no seed. In that case, `R` will create one from the current time and process ID. Hence, different sessions will give different results when random numbers are involved. When you store a workspace and reopen it, you will continue from the seed specified within the workspace. 

---

1. **Fix the random seed to `123`**
```{r}
set.seed(seed = 123)
```
or simply
```{r}
set.seed(123)
```

The random number generator is now fixed to value `123`. We can now ensure that our data generation process is exactly reproducible on every machine using `R` around the world. 

---

#### Drawing random values

---

2. **Draw 10 values from a standard normal distribution and store the values in an object called `draw1`**.
```{r}
draw1 <- rnorm(n = 10)
```
The call `rnorm(n = 10)` results in a draw of 10 values from a standard normal distribution, a distribution with mean zero and variance equal to 1. Drawing from a standard normal distribution is the default in `rnorm` because the default function arguments for `rnorm()` specify:
```{r}
args(rnorm)
```
We can see that by default, values a drawn with `mean = 0` and `sd = 1`. Unless we specify other values for the mean and standard deviation, the defaults will be used. Defaults are a useful property in a developers toolset. We will discuss defaults in more detail later today. 

---

3. **Verify that the means and variances for object `draw1` are indeed conform the default arguments of function `rnorm()`**
```{r}
mean(draw1)
sd(draw1)
```
The means and variances are close, but not equal. However, we've only drawn 10 values from a theoretical distribution. It is statistically not reasonable to expect that a single draw of 10 values will result in an unbiased estimate of the (infinitely large) population parameters.

---

4. **Draw 1000 values from a standard normal distribution and store the values in an object called `draw2`**.
```{r}
draw2 <- rnorm(1000)
```
Note that I do not have to specify `n = 1000`: by default, the ordering of the arguments is used to evaluate a function call. So `rnorm(40, 20, 5)` would generate a draw of 40 values from a normal distribution with mean equal to 20 and a standard deviation of 5. 

---

5. **Determine which object has less bias; `draw1` or `draw2`**
```{r}
means <- c(0, mean(draw1), mean(draw2))
sds <- c(1, sd(draw1), sd(draw2))
n <- c(Inf, 10, 1000)
result <- data.frame(n=n, mean=means, sd=sds)
row.names(result) <- c("population", "draw1", "draw2")
result
```
We have now created a `data.frame` with the results. Perhaps this dataframe becomes easier to read with some rounding:
```{r}
round(result, 3)
```
It is clear that the larger sample size will, in general, yield less bias. However, the bias we obtain is directly dependent on the chosen random seed. For example, if we repeat the above process with `set.seed(125)`, we obtain completely different results:
```{r}
set.seed(125)
draw1b <- rnorm(10)
draw2b <- rnorm(1000)
means <- c(0, mean(draw1b), mean(draw2b))
sds <- c(1, sd(draw1b), sd(draw2b))
result <- data.frame(n=n, mean=means, sd=sds)
row.names(result) <- c("population", "draw1 (seed=125)", "draw2 (seed=125)")
round(result, 3)
```
Now `draw2` has a larger absolute bias from the mean, despite its larger sample size. Note that I renamed the objects to `draw1b` and `draw2b` because otherwise I would overwrite `draw1` and `draw2`. Overwriting these objects would be inefficient, because we will use these objects later on in this exercise.  If, by accident we would have overwritten these objects, we can simply obtain them again by re-fixing the random seed and parsing the exact same function calls. Also note that I do not recreate object `n` as it has not changed. I can simply re-use the object `n` that is already in our `Global Environment`. 

---

The lesson so far is: When you fix your random seed, do not forget that your results are dependent on this random seed. It may be the case that results you obtain are a mere fluke: extremely rare, but obtained because you use a random seed. It would be nice to automate any random number process with different seeds to obtain the sampling distribution of estimates. This we will do on Friday when we consider Monte Carlo simulation. 

---

In the previous exercise we have changed the random seed to `125`. 

---

6. **Re-fix the random seed to `123`**
```{r}
set.seed(123)
```

---

7. **Draw objects `draw1` and `draw2` again, but name them `draw3` and `draw4`, respectively.**
```{r}
draw3 <- rnorm(10)
draw4 <- rnorm(1000)
```

---

8. **Draw object `draw5`: 22 values from a normal distribution with `mean = 8` and `variance = 144`.**
```{r}
draw5 <- rnorm(22, 8, sqrt(144)) #square root of 144 
```
Function `rnorm()` takes the standard deviation $\sigma$, not the variance $\sigma^2$. Hence we take the square root of the variance to obtain the standard deviation cf.

\[\sigma=\sqrt{\sigma^2} = \sqrt{144} = 12\]

---

9. **Verify if object `draw1` is equal to `draw3` and if object `draw2` is equal to `draw4`.**
```{r}
all.equal(draw1, draw3)
all.equal(draw2, draw4)
```
The objects are indeed equal because they originated in an identical order since the same random seed was specified. In other words, `draw1` and `draw3` were generated in random step 1 since `set.seed(123)` and `draw2` and `draw4` were both generated in step 2 since `set.seed(123)`. If we would have reversed the order in which these objects were generated, then the objects would not be equal. For example:
```{r}
set.seed(123)
draw4 <- rnorm(1000)
draw3 <- rnorm(10)
all.equal(draw1, draw3)
all.equal(draw2, draw4)
```
Because the objects are not equal, mean differences are automatically printed by function `all.equal()`.

**Fun fact:** the first 10 cases in `draw1` and `draw4` are equal:
```{r}
cbind(draw1[1:10], draw4[1:10])
```
and if we recreate `draw5`, the results are also equal. 
```{r}
draw5b <- rnorm(22, 8, sqrt(144))
all.equal(draw5, draw5b)
```
This is because these values and objects follow the same steps since `set.seed(123)`. The first 10 cases in `draw1` and `draw4` both stem from the first step since `set.seed(123)` and generating `draw5` and `draw5b` has been the third random step since `set.seed(123)` in both random chains. 

---

10. **Draw 1000 values from the following distributions and inspect them with `summary()`**

- a normal distribution with `mean = 50` and `sd = 20`, 
- a t-distribution with `df = 11` degrees of freedom,
- a uniform distribution with minimum `min = 5` and maximum `max = 6`,
- a binomial distribution with `size = 1` trials and `prob = .8` probablity if success on each trial,
- an F-distribution with `df1 = 1` and `df2 = 2` degrees of freedom,
- a poisson distribution with `lambda = 5` as the mean. 

First, 
```{r}
draw.norm <- rnorm(1000, mean = 50, sd = 20)
draw.t <- rt(1000, df = 11)
draw.unif <- runif(1000, min = 5, max = 6)
draw.binom <- rbinom(1000, size = 1, prob = .8)
draw.F <- rf(1000, df1 = 1, df2 = 2)
draw.pois <- rpois(1000, lambda = 5)
```
It may ease interpretation to create a `data.frame` from these results and call summary on the data frame. 
```{r}
data <- data.frame(normal = draw.norm, 
                   t = draw.t, 
                   uniform = draw.unif, 
                   binomial = draw.binom,
                   Fdistr = draw.F,
                   poisson = draw.pois)
summary(data)
```

---

11. **Plot histograms for the generated values from exercise 10 to inspect the shape of the respective distributions. Use `breaks = 25` for the number of breakpoints in the histogram **
```{r}
hist(draw.norm, breaks = 25)
hist(draw.t, breaks = 25)
hist(draw.unif, breaks = 25)
hist(draw.binom, breaks = 25)
hist(draw.F, breaks = 25)
hist(draw.pois, breaks = 25)
```

---

Alternatively, we can automate this procedure with a simple single line call:
```{r}
invisible(apply(data, 2, function(x) hist(x, breaks = 25)))
```

I use the `invisible()` function to omit the printing of the histogram data and only printing the histogram itself. 

We will learn all about `apply()` in the next meeting for today. 

---

We can use random number generation to simulate data. For example, we can use `sample()` to simulate rolling a single die 100 times:
```{r}
die.roll <- sample(1:6, size = 100, replace = TRUE)
```
We have to set `replace = TRUE` to sample with replacement because the sample size `size = 100` is larger than the set `1:6 = ``r 1:6`. 

---

12.  **Calculate the probability for obtaining each side of the die in `die.roll`. **
```{r}
die.roll.tbl <- table(die.roll)
prop.table(die.roll.tbl)
```
We see that in this try, the values 1 and 2 have a much higher probability than the other values. There results are obtained by first creating a frequency table out of `die.roll`:
```{r}
table(die.roll)
```
and then calculating the probability of the frequencies, given the total number of rolls
```{r}
prop.table(table(die.roll))
```

---

13. **Simulate two rolling dice and calculate the probability of obtaining two sixes over 1500 trials**
```{r}
possible <- rep(1:6, rep(6, 6)) + rep(1:6, 6)
possible
dice.roll <- sample(possible, size = 1500, replace = TRUE)
dice.roll.tbl <-  table(dice.roll) 
prop.table(dice.roll.tbl)
```
The probability equals `r mean(dice.roll == 12)`, which can simply be obtained by calculating the mean over the logical evaluation (i.e. number of `TRUE`'s given the total sample size `1500`)
```{r}
mean(dice.roll == 12)
```


---

14. **Replicate the following empirical example by drawing from a *theoretical* distribution: 14563 students made an exam, but only 3589 successfully passed:**

One way to do this is with `rbinom()`
```{r}
prob.pass <- 3589/14563
passfail <- rbinom(14563, size = 1, prob = prob.pass)
table(passfail)
```
However, we can also obtain an approximation with `sample()`
```{r}
passfail2 <- sample(c("pass", "fail"), 
                    14563, 
                    prob = c(prob.pass, 1-prob.pass), 
                    replace = TRUE)
table(passfail2)
```

---

So far, we have only considered univariate distributions. Let's generate some multivariate data

---

15. **Draw two variables (aka sets or features) of size `1000` from a multivariate normal distribution, both with mean `5` and variance `1` and make sure that they correlate $\rho = .8$.**

There are two ways to go about this.

---

- Approach 1: generate two univariate normal distributions and *make them correlate*:
```{r}
rho = .8
var1 <- rnorm(1000, 5, sqrt(1)) #first variable
var2 <- scale(var1) * rho + rnorm(1000, 0, sqrt(1 - rho^2)) #second variable step 1
var2 <- var2 * sd(var1) + mean(var1) #second variable step 2
data <- data.frame(var1 = var1, var2 = var2) #combined data frame
colMeans(data) #means
var(data) #variance/covariance matrix
cor(var1, var2) #correlations
```
We see that the means are indeed approximately `5`, the variances approximate `1` and the correlation is approximately `.8`. In the case where the variance equals `1`, the correlation is equal to the covariance. For example, if the variance were exactly `1`, the correlation would be:

\[\rho_{X,Y} = \frac{\text{COV(X,Y)}}{\sigma_X\sigma_Y} = \frac{.7925752}{1\times1} = .7925752.\]

In our case, the correlation equals:

\[\rho_{X,Y} =  \frac{.7925752}{\sqrt{0.9847586}\times\sqrt{.9830075}} = .8055587 \approx .80.\]

---

- Approach 2: generate multivariate data from the multivariate normal distribution:
```{r}
#you need function rmvnorm from package mvtnorm
sigma <- matrix(c(1, .8, .8, 1), nrow = 2, ncol = 2) #variance/covariance matrix
data <- mvtnorm::rmvnorm(n = 1000, sigma = sigma, mean = c(5, 5))
colMeans(data)
var(data)
```
Approach two yields approximately the same result as approach 1. Note that both approaches are equally valid: differences between them are only due to random variation. If we would simulate the processes repeatedly an infinite ($\infty$) number of times, the average parameters over these repeated simulations will converge to the population parameters. 

---

End of Practical
