---
title: "Practical_O"
author: "Gerko Vink"
date: "Statistical Programming in R"
output: html_document
---
  
  ```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We use the following packages:
```{r message=FALSE, warning=FALSE}
library(plyr)
library(dplyr)
library(magrittr)
library(ggplot2)
```

---

#### Exercise

1. **Sample 100 samples from a standard normal distribution. **
2. **For each of these samples, calculate the following statistics for the mean:**

 - absolute bias
 - standard error
 - lower bound of the 95% confidence interval
 - upper bound of the 95% confidence interval
 - a logical indicator, denoting `TRUE` when the population parameter is in the interval
 
3. **Create a plot that demonstrates the following**

    ***"A replication of the procedure that generates a 95% confidence interval 
    that is centered around the sample mean would cover the population value at 
    least 95 out of 100 times"*** [(Neyman, 1934)](https://www.google.nl/url?sa=t&rct=j&q=&esrc=s&source=web&cd=1&ved=0ahUKEwjk4LaD6LrXAhUGC8AKHdATBhMQFggoMAA&url=http%3A%2F%2Fwww.stat.cmu.edu%2F~brian%2F905-2008%2Fpapers%2Fneyman-1934-jrss.pdf&usg=AOvVaw3KRw9DSklPihrYIM0_zRrC)
    
4. **Present a table containing all simulated samples for which the resulting confidence interval does not contain the population value.**

---

End of Practical

