---
title: "Exercise in markdown"
author: "Gerko Vink"
date: "Statistical programming in `R`"
output: html_document
---

---

1. **Create an object a with element (value) `1`**
```{r}
a <- 1
```

We can see that value `1` is assigned to a, but we cannot verify that it is 
indeed stored. 

---

2. **Verify that 1 is stored in `a`**
```{r}
a
```

Pfff. Luckily there is nothing to worry about. The value 1 is indeed stored in
object `a`. Have you noticed that these lines of text do not contain more than
80 characters? That is the conventional rule that has been agreed upon by the
council of R-nerds. Just kidding, but it is somehow convenient. Much more 
convenient than having a line of code that kind of extends beyond the 80 
characters mark. But in Markdown this is not an issue! 

---

3. **Square `a`**
```{r}
a^2
```

Still 1, that's nice!

---

4. **Create `b` and assign `a+a`. Check if `b` is indeed `a+a`. **
```{r}
b <- a + a
b == a + a
```

`True!` R is telling us that the equality we tested is indeed true. In other 
words, everything before the == is equal to that which is posed after the ==. 
Now we are talking. 

---

5. **Square `b`, multiply the answer with $\frac{a}{b}$ and call the result `c`. Then, take the square root of $c^b$ and multiply it with $\frac{a}{b^6}$. **
```{r}
c <- b^2 * (a / b)
sqrt(c^b) * (a / b^6)
```

Nice! But anything beyond a couple of decimals may not be relevant. 

---

6. **Round the answer from the previous exercise to 3 decimals.**
```{r}
round(sqrt(c^b) * (a / b^6), 3)
```

or, alternatively
```{r}
d <- sqrt(c^b) * (a / b^6)
round(d, 3)
```


Do you notice that the first solution is shorter, but the second is much more 
insightful? If you share your code, or plan to use it later on, clarity is of
great importance. A small effort now, saves you hours later. 

Now you know how to use `R` as a calculator and `R-Studio` as a typesetting 
device. 

---

End of exercise