---
title: "Practical B"
author: "Javier Garcia-Bernardo"
date: "Statistical Programming with R"
output: 
   html_document:
    toc: true
    toc_depth: 5
    toc_float: 
      collapsed: false
      smooth_scroll: false
    number_sections: false
    code_folding: hide
---




```{r echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = FALSE, include = FALSE)
```

In the practical I will be explicitly specifying the package of each library (e.g. readr::read_delim). Specifying the package is not required and can be cumbersome for the most commonly used functions. It is however a good idea to understand the source of the functions you are using.

In the following exercises we will use the same dataset as last time, boys.

(if you haven't done the previous practical) You need to download You need to download `dataset_boys.csv` from [here](https://javier.science/R/#day-2-nov-26th) and add it to a folder "data" in the same folder as your markdown file.

Then, read the file dataset_boys.csv using the code in the next cell.

## Reading data

```{r}
#install.packages(c("readr"),repos = "http://cran.us.r-project.org")
library(readr)
#Reading the file
boys <- readr::read_delim("data/dataset_boys.csv",delim=",")
is_tibble(boys)
```


## Exercise 1-4: Tidyverse operations


---

1. Filter: **Create a new tibble What is the mean age of boys younger than 15 years of age that do not live in region `north`?**

```{r}
age_subset <- dplyr::filter(boys, (age < 15), (reg != "north"))
mean(age_subset$age, na.rm=TRUE)
```

2. Select: **Select variables bmi and wgt of the tibble you just created**

```{r}
dplyr::select(age_subset, bmi, wgt)
```

3. Mutate: **Create two new variables, one call hgt_square (htg^2), and another called bmi_2 (10000  wgt/hgt_square)**. Use one command only

```{r}
# The arguments are evaluated sequentially in tidyverse, by the time we get to bmi_2 the computer has already created hgt_square
dplyr::mutate(boys, 
              hgt_square = htg^2,
              bmi_2 = 10000 * wgt / hgt_square)
```

4. **group_by + summarize: Group by region, and calculate the average bmi, wgt and hgt. Use a pipe if you dare**

```{r}
# Without a pipe
gr_boys <- dplyr::group_by(boys, reg)  
dplyr::summarize(gr_boys, mean_bmi = mean(bmi, na.rm=TRUE),
                   mean_wgt = mean(wgt, na.rm=TRUE),
                   mean_hgt = mean(hgt, na.rm=TRUE))

# With a pipe
# The arguments are evaluated sequentially in tidyverse, by the time we get to bmi_2 the computer has already created hgt_square
boys %>% 
  dplyr::group_by(reg) %>% 
  dplyr::summarize(mean_bmi = mean(bmi, na.rm=TRUE),
                   mean_wgt = mean(wgt, na.rm=TRUE),
                   mean_hgt = mean(hgt, na.rm=TRUE))


```

4. **arrange: Sort the dataset by bmi, what can you say about the boys with low bmi?**

```{r}
#The boys are small babies
dplyr::arrange(boys, bmi)
```

## Exercises 5--8: Data transformations
5. **Imagine you have the dataset tidyr::table2, could you easily create the cases of TB per capita?** (you only need to print `tidyr::table2`)

```{r}
tidyr::table2
```

6. **Transform the data to make it tidy, and calculate the cases of TB per 1 million people? Which country had the highest cases per capita?** Do you have values in the column names and you need to make the data longer? Or do you have more than one variable in one column and you need to make the data wider?

```{r}
# With a pipe (but you could do it with intermediate steps as well)
table2 %>%  
  tidyr::pivot_wider(id_cols = c("country","year"),
                    names_from = "type",
                    values_from = "count") %>% 
  dplyr::mutate(cases_per_capita = 1E6*cases/population) %>% 
  dplyr::arrange(desc(cases_per_capita))
```

7. **Tidy the datasets tidyr::table4a (cases of TB) and tidyr::table4b (population). Save the output table with names table4a_tidy and table4ab_tidy** Do you have values in the column names and you need to make the data longer? Or do you have more than one variable in one column and you need to make the data wider?

```{r}
table4a_tidy <- tidyr::pivot_longer(table4a, cols = c("1999","2000"),
                                names_to = "year",
                                values_to = "TB") 
head(table4a_tidy)

table4b_tidy <-  tidyr::pivot_longer(table4b, cols = c("1999","2000"),
                                   names_to = "year",
                                    values_to = "Pop") 
head(table4b_tidy)
```

8. **Join table4a_tidy and table4b_tidy, and calculate the cases of TB per 1 million people? Which country had the highest cases per capita?** Could you have done this with table4a and table4b?

```{r}
#With a pipe
dplyr::inner_join(table4a_tidy, table4b_tidy, on = c("country", "year")) %>% 
    dplyr::mutate(cases_per_capita = 1E6*TB/Pop) %>% 
    dplyr::arrange(desc(cases_per_capita))
  
```

## Exercises 9--11: Linear regression

9. **Using the boys dataset, fit a model where the weight depends on the age and the region.** 

```{r}
fit <- stats::lm(wgt ~ age + reg, data=boys)
fit
```

10. **Print the regression table using the function summary. Does the region have a significant effect?** 

```{r}
summary(fit)
```

11. **Check the assumptions of linear regression using plot(). Are the assumptions met? What are the biggest outliers? (find them in the data)** Homoscedasticity and independence of residuals (which = 1). Normality of residuals (which = 2 and which = 5). 

```{r}
plot(fit, which = 1)
plot(fit, which = 2)
plot(fit, which = 6)
boys[c(733,610,574),]
```


## Exercise 12. Saving data

12. **Save the data `boys` as an RDS file, and as an Excel file**. Remember to save your data to the data folder!

```{r}
readr::write_rds(boys, "data/boys.rds")
writexl::write_xlsx(boys, "data/boys.xlsx")
```



## Exercise 13--14. Some visualizations


Tidyverse is great because it connects very well with ggplot. You'll learn more next time, but let's demonstrate using the datasaurus dataset. There are 12 datasets composed of a `x` variable and a `y` variable, created in a way that mean(x), mean(y), sd(x), sd(y).

First, install the library datasauRus and load it


```{r}
#install.packages("datasauRus")
library(datasauRus)
```


13. Group datasauRus::datasaurus_dozen by the variable `dataset` and calculate the mean and standard deviation of both x and y, and the correlation (function `cor()`) between x and y.

```{r}
datasauRus::datasaurus_dozen %>% 
      dplyr::group_by(dataset) %>% 
      dplyr::summarize(mean_x = mean(x),
                mean_y = mean(y),
                sd_x = sd(x),
                sd_y = sd(y),
                cor(x,y))
```

14. **Use the following code to visualize the data.**

```{r, class.source = 'fold-show'}
ggplot(datasaurus_dozen, aes(x=x, y=y, colour=dataset))+
    geom_point()+
    theme(legend.position = "none")+
    facet_wrap(~dataset, ncol=3)
```




End of Practical

---

#### References


Matejka, J., & Fitzmaurice, G. (2017). Same Stats, Different Graphs: Generating Datasets with Varied Appearance and Identical Statistics through Simulated Annealing. CHI 2017 Conference proceedings: ACM SIGCHI Conference on Human Factors in Computing Systems.




