---
title: "Practical H"
author: "Gerko Vink"
date: "Statistical Programming in R"
output: html_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
printAnswers <- TRUE
```

We use the following packages:
```{r}
library(ggplot2)
library(dplyr)
```

---

#### Exercises

---

Again, just like last time it is wise to start with fixing the random seed.

```{r}
set.seed(123)
```

---

1. **Generate two random samples of 10 numbers from a normal distribution with the below specifications. Test the null hypothesis that the population mean is 0.**

- $\mu = 0$ and $\sigma = 2$
- $\mu = 1.5$ and $\sigma = 2$

```{r eval=printAnswers, echo=printAnswers}
x <- rnorm(10, mean = 0, sd = 2)
t.test(x)

x <- rnorm(10, 1.5, 2)
t.test(x)
```

---

2. **Write a function that generates a random sample of `n` numbers from a normal distribution with a user defined mean (i.e. a mean that you can choose when running the function) and standard deviation 1, and returns the `p.value` for the test that the mean is 0.**

```{r eval=printAnswers, echo=printAnswers}
p.value.t <- function (n, mu) {
  x <- rnorm(n, mu, 1)
  t.test(x)$p.value
}

p.value.t(n = 30, mu = 3)
```

---


3. **Use the function of Exercise 3 to generate 50 $p$-values with $n=10,\mu=0$, and make a `qqplot` to compare distribution of the $p$-values with a uniform $[0,1]$ variable.**

```{r eval=printAnswers, echo=printAnswers}
y <- numeric(50)
for (i in 1:50) {
  y[i] <- p.value.t(n = 10, mu = 0)
}

qqplot(x=qunif(ppoints(50)), y)
```

The p-values follow a uniform distribution. 

---

In a study that examined the use of acupuncture to treat migraine headaches, consenting patients on a waiting list for treatment for migraine were randomly assigned in a 2:1:1 ratio to acupuncture treatment, a "sham" acupuncture treatment in which needles were inserted at non-acupuncture points, and waiting-list patients whose only treatment was self-administered (Linde et al., 2005). The "sham" acupuncture treatment was described to trial participants as an acupuncture treatment that did not follow the principles of Chinese medicine. 

---

4. **What is the conclusion when the outcome is classified according to numbers of patients who experienced a greater than 50% reduction in headaches over a four-week period, relative to a pre-randomization baseline?**

Use the following data
```{r}
data <- matrix(c(74, 71, 43, 38, 11, 65), nrow = 2, ncol = 3)
colnames(data) <- c("Acupuncture", "Sham", "Waiting list")
rownames(data) <- c("> 50% reduction", "< 50% reduction")
data
```

We start with calculating the $X^2$-test:
```{r eval=TRUE, echo=TRUE}

X2test <- 
  data %>%
  chisq.test()

X2test
```

which is extremely significant. We can then calculate the expected cell frequencies
```{r}
X2test$expected
```
and the raw residual
```{r}
X2test$observed - X2test$expected
```

as well as the Pearson residual
```{r}
X2test$residuals
```
to infer the difference in observed and expected cell frequencies. Patients on the waiting list experience `> 50% reduction` much less than we would expect under independence of treatment and outcome.

---

5. **Patients who received the acupuncture and sham acupuncture treatments were asked to guess their treatment at the end of their trial. What would you conclude from this data?**

```{r}
data <- matrix(c(82, 17, 30, 30, 26, 16), nrow = 3, ncol = 2)
colnames(data) <- c("Acupuncture", "Sham")
rownames(data) <- c("Chinese", "Other", "Don't know")
data
```

We again start with calculating the $X^2$-test:
```{r eval=TRUE, echo=TRUE}

X2test <- 
  data %>%
  chisq.test()

X2test
```

which is very significant. We can then calculate the expected cell frequencies
```{r}
X2test$expected
```
and the raw residual
```{r}
X2test$observed - X2test$expected
```

as well as the Pearson residual
```{r}
X2test$residuals
```

We find that people who are receiving *true* `Acupuncture` are more inclined to believe that they receive `Chinese` acupuncture than we would expect under independence, while people wo received `Sham` acupuncture are more inclined to believe that they receive `Other` type of acupuncture. `Don't know` is more or less similarly distributed over the observed and expected frequencies.

---

In the following simulation experiment we investigate least-squares estimation of the mean.

---

6. **Start by drawing a 100 values from a normal distribution with $\mu = 3$ and $\sigma = 7$. Use seed value `32083`.**


```{r}
set.seed(32083)

x <- rnorm(100, mean=3, sd=7)
```

---

7. **Next, confirm that the sample mean of the values in x is near 4.166.**
```{r}
mean(x)
```

---

8. **Calculate the sample mean's sum of squared deviations from $\mu$. The sum of squared deviations from mu is defined as: **
$$ \sum_{i=1}^{100} (x_i - \mu)^2,$$
There is a slow way
```{r}
mu = 3
summed <- x-mu
sum.sq <- sum(summed^2)
```

And a fast way
```{r}
sum.sq2 <- apply(outer(x, mu, "-")^2, 2, sum)
```
Both solutions are identical
```{r}
identical(sum.sq, sum.sq2)
```

---

9. **Now create a function that automates the calculation of the sum of squares for any given $\mu$. Call the function `lsfun` because we are going to identify the least squares estimate in exercise 8.**

```{r eval=printAnswers}
lsfun <- function(meanest) apply(outer(x, meanest, "-")^2, 2, sum)
```

or, 100% equivalently, but easier to spot as a function:
```{r}
lsfun <- function(meanest){
  apply(outer(x, meanest, "-")^2, 2, sum)
}
```

---

10. **Plot the curve of your least square function such that you can identify the minimum of the curve (i.e. the location for $x$ where the sum of the squared deviations are the lowest).**
```{r}
curve(lsfun, from = 4.16, to = 4.17)
```

---


11. **Repeat the experiment from 10 with the following $X \sim \mathcal{N}(\mu, \sigma^2)$ normal samples of length $n=100$, but now use the sample mean $\bar{x}$ in your function `lsfun()`. Let the function plot the curve and print the location where the minimum of the sum of the squares is located each time. Fix the seed to `set.seed(123)`:**

- $x \sim \mathcal{N}(3, 7)$
- $x \sim \mathcal{N}(15, 12)$
- $x \sim \mathcal{N}(0, 2)$
- $\sqrt{x} \sim \mathcal{N}(0, 2)$, i.e. you need to square $x$

Hint: use the sample mean $\bar{x}$ as the center of your graph and add/subtract e.g. `.5` from this value to plot a range.

First, we fix the random seed
```{r}
set.seed(123)
```

Code-wise it is efficient to write a function that does the repetitive experimentation. That way we have to write some lines of code only once. 
```{r}
plotfun <- function(x, meanest, plot = TRUE){
  xbar <- mean(x)
  lsfun <- function(meanest){ apply(outer(xbar, meanest, "-")^2, 2, sum)}
  # lsfun <- function(mu, xbar) {
  #   summed <- x-mu
  #   sum.sq <- sum(summed^2)
  #   return(sum.sq)
  # }
  if (plot) {
    curve(lsfun, from = xbar - .5, to = xbar + .5, 
          ylab=expression(paste(Sigma," ",e^2))) #nicer y-axis label
  }
  return(cat("The mean is:", xbar, "\n"))
}
```

- For $x \sim \mathcal{N}(3, 7)$:
```{r eval=printAnswers, echo=printAnswers}
rnorm(100, mean=3, sd=sqrt(7)) %>%
  plotfun(meanest = 3, plot = TRUE)
```

- For $x \sim \mathcal{N}(15, 12)$
```{r eval=printAnswers, echo=printAnswers}
rnorm(100, mean=15, sd=sqrt(12)) %>%
  plotfun(meanest = 15, plot = TRUE)
```

- For $x \sim \mathcal{N}(0, 2)$:
```{r eval=printAnswers, echo=printAnswers}
rnorm(100, mean=0, sd=2) %>%
  plotfun(meanest = 0, plot = TRUE)
```

- For $\sqrt{x} \sim \mathcal{N}(0, 2)$:
```{r eval=printAnswers, echo=printAnswers}
rnorm(100, mean=0, sd=2)^2 %>%
  plotfun(meanest = 0, plot = TRUE)
```


---

End of practical. 
